XOr-Some Bayesian Power analysis

### 2021-09-17 analysis/power/xor-some_bayesian-power.Rmd
copy over helper functions, add first stab at REs

## Simulate initial data
We assume an effect size of 0.15 for the predictors prior, competence and relevance, and, therefore, simulate data as coming from a Normal distribution with $\mu=±.15$ and $\sigma=.05$ (because this $\sigma$ would give us app. 95% of the data in the region we defined as a positive / negative effect). The target is simulated via $\mathcal{N}(\mu_t, 1)$, with $mu_t = \beta_0 + \beta_{pri} + \beta_{comp} + \beta_{rel} + \beta_{trigger} + interactions$. The coefficients for interactions and the trigger are assumed to be 0.

From each participant, we get four data points for xor and four data points for some, one data point per condition (prior X competence X relevance, mapped to triggers at random). For each condition (for each trigger), there are four stories for which the random intercept may vary.

@@

# simulate toy data set for one subject 
d_init <-
  # create the 8 conditions
  tibble(prior = rep(c(0, 1), each = 4),
         comp = rep(c(0, 0, 1, 1), times = 2),
         rel = rep(c(0,1), times = 4),
         main_type = sample(rep(c("xor", "some"), times=4))) %>% 
  rowwise() %>%
  # sample effects based on condition
  mutate(
    subj_ID = 1, 
    intercept = 0,
    prior_obs = ifelse(prior == 0, rnorm(1, sd = 0.05), 
                       rnorm(1, -0.15, 0.05)),
    comp_obs = ifelse(comp == 0, rnorm(1, sd = 0.05), 
                       rnorm(1, 0.15, 0.05)),
    rel_obs = ifelse(rel == 0, rnorm(1, sd = 0.05), 
                       rnorm(1, 0.15, 0.05)),
    # sample response
    # what do we do about the interactions? can they safely be ignored?
    target = rnorm(1, 
                   mean = (intercept + prior_obs + comp_obs + rel_obs), 
                   sd = 1) # how to set / detrmine this properly?
  )
d_init
```

@@

```{r, eval=FALSE}
# TODO
# just old stuff below
get_new_data <- function(d, N) {
    if(N == 0) {
      data <- d
    } else if (N > 47) {
      data <- add_predicted_draws(model=pilot_model, 
                                     newdata = pilot_data %>% 
                                       filter(workerid %in% sample(unique(pilot_data$workerid), N, replace = T)),
                                     n = 1) %>% 
                 mutate(workerid = paste(workerid, letters[1], sep = "_")) %>%
                   rbind(., d)
    } else {
      data <- add_predicted_draws(model=pilot_model, 
                                     newdata = pilot_data %>% 
                                       filter(workerid %in% sample(unique(pilot_data$workerid), N, replace = F)),
                                     n = 1) %>% 
                 mutate(workerid = paste(workerid, letters[1], sep = "_")) %>%
                   rbind(., d)
    }

@@

## Extract contrasts of interest

```{r}
# TODO
```

@@

```{r}
# TODO
```

>>>>

## Simulate initial data
We assume an effect size of 0.15 for the predictors prior, competence and relevance, and, therefore, simulate data as coming from a Normal distribution with $\mu=±.15$ and $\sigma=.05$ (because this $\sigma$ would give us app. 95% of the data in the region we defined as a positive / negative effect).

The target response variable is simulated via $\mathcal{N}(\mu_t, 1)$, with $mu_t = \beta_0 + \beta_{pri} + \beta_{comp} + \beta_{rel} + \beta_{trigger} + interactions$. The coefficients for interactions and the trigger effect are assumed to be 0.

From each participant, we get four data points for xor and four data points for some, one data point per condition (prior X competence X relevance, mapped to triggers at random). For each condition (and for each trigger), there are four stories for which the random intercept may vary.

@@

# simulate toy data set for one subject 
d_init <-
  # create the 8 conditions
  tibble(prior = rep(c(0, 1), each = 4),
         comp = rep(c(0, 0, 1, 1), times = 2),
         rel = rep(c(0,1), times = 4),
         main_type = as.factor(sample(rep(c("xor", "some"), times=4)))) %>% 
  rowwise() %>%
  # sample effects based on condition
  mutate(
    subj_ID = 1, 
    intercept = 0,
    prior_obs = ifelse(prior == 0, rnorm(1, sd = 0.05), 
                       rnorm(1, -0.15, 0.05)),
    comp_obs = ifelse(comp == 0, rnorm(1, sd = 0.05), 
                       rnorm(1, 0.15, 0.05)),
    rel_obs = ifelse(rel == 0, rnorm(1, sd = 0.05), 
                       rnorm(1, 0.15, 0.05)),
    # sample response
    # what do we do about the interactions? can they safely be ignored?
    target = rnorm(1, 
                   mean = (intercept + prior_obs + comp_obs + rel_obs), 
                   sd = 1) # how to set / determine this properly?
  )
d_init
```

@@

First, add random effects to the simulation based on random effects estimated on pilot data.
```{r, eval=FALSE, warnings=FALSE, message=FALSE}
# fit model on pilot data to get random effects estimates
pilot1 <- read_csv("./../data/pilots/pilot2_critical_zScore_wide_tidy.csv") %>%
  select(-competence_wUtt, -relevance_wUtt)
pilot2 <- read_csv("./../data/pilots/pilot3_critical_zScore_wide_tidy.csv")

d_both <- pilot1 %>% mutate(pilot = 1) %>% 
  rbind(., pilot2 %>% mutate(pilot = 2)
        )
# Full model 
d_both <- d_both %>% mutate(
  main_type = as.factor(main_type)
)
# some = 0, xor = 1
contrasts(d_both$main_type) 
# try to fit maximal model with brm
model_SI <- brm(target ~ prior*competence*relevance* main_type +
                   (1 + prior + competence + relevance + main_type || submission_id) +
                   (1 | title),
                 data = d_both,
                 control = list(adapt_delta = 0.95),
                 cores = 4,
                 iter = 3000)

# Get the by-story estimates
# check if this code can be less ugly
story_rand_int <- c(ranef(model_SI)[['title']][1:(length(ranef(model_SI)[['title']])/4)])

# get by-subj random intercepts and slopes
# TODO

# idea: sample from the vectors of observed estimates when simulating 
# (across predictors (slopes for pri vs rel vs comp for subjects) 
# and story types (comp X pri X rel condition of story))
#### --> Is this okay? 
```

Build helper function for simulating data for a given number of subjects.
[stuff from here on is not finished yet]
```{r, eval=FALSE}
# TOFINISH
# N = number of subjects to simulate
get_new_data <- function(d, N) { # d probably unnecessary

      data <-
  # create the 8 conditions, manually, for double-checking
        tibble(prior = rep(c(0, 0, 0, 0, 1, 1, 1, 1), times = N),
         comp = rep(c(0, 0, 1, 1, 0, 0, 1, 1), times = N),
         rel = rep(c(0, 1, 0, 1, 0, 1, 0, 1), times = N),
         main_type = rep(sample(c("xor", "some", "xor", "some", "xor", "some", "xor", "some")), times=N),
         # assign subject ID
         subj_ID = rep(1:N, each = 8),
         # prepare parameters 
         intercept = 0,
         # sample random effects
         story_int = sample(story_rand_int, N*8, replace = T)
         # TODO: analogously sample by-subj effects
         ) %>% 
  rowwise() %>%
  # sample effects based on condition
  mutate(
    prior_obs = ifelse(prior == 0, rnorm(1, sd = 0.05), 
                       rnorm(1, -0.15, 0.05)),
    comp_obs = ifelse(comp == 0, rnorm(1, sd = 0.05), 
                       rnorm(1, 0.15, 0.05)),
    rel_obs = ifelse(rel == 0, rnorm(1, sd = 0.05), 
                       rnorm(1, 0.15, 0.05)),
    # sample response
    # what do we do about the interactions? can they safely be ignored?
    target = rnorm(1, 
                   mean = (intercept + story_int + prior_obs + comp_obs + rel_obs), 
                   sd = 1) # how to set / determine this properly?
  )

@@

Build helper function for simulating and computing regression model for k iterations (tracked as seeds). Also extract contrasts of interest here.
```{r, eval=FALSE}
# TODO:
# just copied over old stuff below

# simulate data and update
sim_data_fit <- function(seed, N) {
  set.seed(seed)

  # possibly add more participants
  data <- get_new_data(d, N)

  # mutate main_type to factor

  # update model fit with new data
  update(model_init,
         newdata = data,
         seed = seed) %>% 
    # extract posterior draws
    spread_draws(b_Intercept, b_syntax_dev1, b_trial_dev1, `b_syntax_dev1:trial_dev1`) %>%
    # extract contrasts of interest, especially effect of syntax by-trial 
  mutate(critical_subj = b_Intercept + b_syntax_dev1 - b_trial_dev1 - `b_syntax_dev1:trial_dev1`,
         critical_pred = b_Intercept - b_syntax_dev1 - b_trial_dev1 + `b_syntax_dev1:trial_dev1`,
         syntax_critical = critical_subj - critical_pred, # subject vs pred syntax
         filler_subj = b_Intercept + b_syntax_dev1 + b_trial_dev1 + `b_syntax_dev1:trial_dev1`,
         filler_pred = b_Intercept - b_syntax_dev1 + b_trial_dev1 - `b_syntax_dev1:trial_dev1`,
         syntax_filler = filler_subj - filler_pred) %>% # subject vs predicate syntax
  select(b_Intercept, b_syntax_dev1, b_trial_dev1, `b_syntax_dev1:trial_dev1`, critical_subj, critical_pred, syntax_critical, filler_subj, filler_pred, syntax_filler) %>%
  gather(key, val) %>%
  group_by(key) %>%
    # compute summary statistics 
  summarise( 
    mean = mean(val),
    lower = quantile(val, probs = 0.025),
    upper = quantile(val, probs = 0.975)
  )
##########  
  # use stuff below here
  spread_draws(b_Intercept, b_prior, b_competence, b_relevance,
                          b_main_typexor, `b_prior:competence`,
                          `b_prior:relevance`, `b_competence:relevance`,
                          `b_prior:main_typexor`, `b_competence:main_typexor`,
                          `b_relevance:main_typexor`, `b_prior:competence:relevance`, 
                          `b_prior:competence:main_typexor`, `b_prior:relevance:main_typexor`,
                          `b_competence:relevance:main_typexor`, 
                          `b_prior:competence:relevance:main_typexor`) %>% 
  mutate(
    prior_xor = b_prior + `b_prior:main_typexor`,
    prior_some = b_prior,
    competence_xor = b_competence + `b_competence:main_typexor`,
    competence_some = b_competence,
    relevance_xor =  b_relevance + `b_relevance:main_typexor`,
    relevance_some = b_relevance
  ) -> model_SI_posteriors

# check P that the effects are positive / negative / no effect present
posterior_hypotheses <- model_SI_posteriors %>% 
  select(prior_xor, prior_some, 
         competence_xor, competence_some,
         relevance_xor, relevance_some) %>%
  gather(key, val) %>%
  group_by(key) %>% mutate(positive = mean(val > 0.05),
                           negative = mean(val < -0.05),
                           no = mean(val %>% between(-0.05, 0.05))) %>%
  summarise(positive_eff = mean(positive),
            negative_eff = mean(negative),
            no_eff = mean(no))
}
```

@@

Run the simulations and test the hypothesis of interest for each iteration and compute the proportion of successful runs.
```{r, eval=FALSE}
# TODO
test_conjunction_of_all_hypotheses <-  function(posterior_hypotheses) {
  posterior_hypotheses %>% 
    mutate(hypothesis_true = case_when(
      key == 'competence_some' ~ positive_eff > 0.95,
      key == 'competence_xor'  ~ positive_eff > 0.95,
      key == 'prior_some' ~ negative_eff > 0.95,
      key == 'prior_xor'  ~ negative_eff > 0.95, 
      key == 'relevance_some' ~ positive_eff > 0.95,
      key == 'relevance_xor'  ~ positive_eff > 0.95
    )) %>% 
    pull(hypothesis_true) %>% all()
}

# again old stuff below

# add iterating over Ns
sim1 <-
  tibble(seed = 1:100) %>% 
  mutate(tidy = map(seed, sim_data_fit, 23)) %>% 
  unnest(tidy)

# map each iter to test_conjunction_of_all_hypotheses
sim1 %>%
  filter(key == "syntax_critical") %>%
  mutate(check_syntax = ifelse(lower > 0, 1, 0)) %>%
  summarise(power_syntax = mean(check_syntax))
